Linting Linearity in Core/System FC

### 2023-01-19 report.tex
Linear types section

\begin{itemize}
    \item Linear types were only added to the language roughly \emph{31 years
        after} Haskell's inception, unlike Rust and Idris2 which were
        designed with linearity from the start. It is an especially difficult
        endeavour to add linear types to a well-established language with a
        large library ecosystem and many active users, rather than to develop
        the language from the ground up with linear types in mind, and the
        successful approach as implemented in GHC 9.0, the leading Haskell
        compiler, was based on Linear Haskell\cite{linearhaskell}, where a
        linear type system designed with retaining backwards-compatibility and
        allowing code reuse across linear and non-linear users of the same
        libraries in mind was described. We describe Linear Haskell in detail in
        section~\ref{linear-haskell-section}.

@@

In order to further give well defined semantics to a linear type system, we
present a linearly typed lambda calculus, a very simple language with linear
types, by defining what are syntatically valid programs through the grammar and
what programs are well typed through the typing rules.

An expression in the linearly typed lambda calculus can be $\star$...
A type in the linearly typed lambda calculus can be $\star$...

\begin{figure}[h]
\[
\begin{array}{c}
  \begin{array}{lcl}
    A,B & ::= & \star\\
        & \mid & A \lolli B\\
        & \mid & A \oplus B\\
        & \mid & A \tensor B\\
        & \mid & A \with B\\
        & \mid & \bang A\\
   \end{array}
\qquad
  \begin{array}{lcl}
      M,N & ::= & \star \mid \llet{\star = M}{N} \\
        & \mid & u \\
        & \mid & \lambda u. M \mid M~N\\
        & \mid & \textrm{inl}~M \mid \textrm{inr}~M \mid
        \ccase{M}{\textrm{inl}~u_1 \to N_1;\textrm{inr}~u_2 \to N_2}\\
        & \mid & M \tensor N \mid \llet{u_1 \tensor u_2 = M}{N}\\
        & \mid & M \with N \mid \textrm{fst}~M \mid \textrm{snd}~M \\
        & \mid & \bang M \mid \llet{!u = M}{N} \\
   \end{array}
\end{array}
\]
\caption{Grammar for a linearly-typed lambda calculus}
\end{figure}

\begin{figure}[h]
\[
  \begin{array}{c}
    \infer*[right=($\tensor I$)]
    {\Gamma ; \Delta \vdash M : A \and \Gamma ; \Delta' \vdash N : B}
    {\Gamma ; \Delta , \Delta' \vdash (M \tensor N) : A \tensor B}
\qquad
    \infer*[right=($\tensor E$)]
    {\Gamma ; \Delta \vdash M : A \tensor B \and \Gamma ; \Delta', u{:}A, v{:}B \vdash N : C}
    {\Gamma ; \Delta , \Delta' \vdash \llet{u \tensor v}{N} : C}
\\[1em]
    \infer*[right=($\lolli I$)]
    {\Gamma ; \Delta , u{:}A \vdash M : B}
    {\Gamma ; \Delta \vdash \lambda u. M : A \lolli B}
\qquad
    \infer*[right=($\lolli E$)]
    {\Gamma ; \Delta \vdash M : A \lolli B \and \Gamma ; \Delta' \vdash N : A}
    {\Gamma ; \Delta, \Delta' \vdash M~N : B}
\\[1em]
    \infer*[right=($\with I$)]
    {\Gamma ; \Delta \vdash M : A \and \Gamma ; \Delta \vdash N : B}
    {\Gamma ; \Delta \vdash M \with N : A \with B}
    \qquad
    \infer*[right=($\with E_L$)]
    {\Gamma ; \Delta \vdash M : A \with B}
    {\Gamma ; \Delta \vdash \textrm{fst}~M : A}
\qquad
    \infer*[right=($\with E_R$)]
    {\Gamma ; \Delta \vdash M : A \with B}
    {\Gamma ; \Delta \vdash \textrm{snd}~M : B}
\\[1em]
    \infer*[right=($\oplus I_L$)]
    {\Gamma ; \Delta \vdash M : A}
    {\Gamma ; \Delta \vdash \textrm{inl}~M : A \oplus B}
\qquad
    \infer*[right=($\oplus I_R$)]
    {\Gamma ; \Delta \vdash M : B}
    {\Gamma ; \Delta \vdash \textrm{inr}~M : A \oplus B}
\\[1em]
    \infer*[right=($\oplus E$)]
    {\Gamma ; \Delta \vdash M : A \oplus B \and \Gamma ; \Delta', w_1{:}A \vdash N_1 : C \and
    \Gamma ; \Delta', w_2{:}B \vdash N_2 : C}
    {\Gamma ; \Delta, \Delta' \vdash \ccase{M}{\textrm{inl}~w_1 \to N_1 \mid \textrm{inr}~w_2 \to N_2} : C}
    \\[1em]
    \infer*[right=($\star I$)]
    { }
    {\Gamma ; \cdot \vdash \star : \star}
\quad    \infer*[right=($\star E$)]
    {\Gamma ; \Delta \vdash M : \star \and \Gamma ; \Delta' \vdash N : B}
    {\Gamma ; \Delta, \Delta' \vdash \llet{\star = M}{N} : B}
\\[1em]
    \infer*[right=($\bang I$)]
    {\Gamma ; \cdot \vdash M : A}
    {\Gamma ; \cdot \vdash !M : !A}
\qquad    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta \vdash M : \bang A \and \Gamma, u{:}A ; \Delta' \vdash N : C}
    {\Gamma ; \Delta, \Delta' \vdash \llet{!u = M}{N} : C}
  \end{array}
\]
\caption{Typing rules for a linearly-typed lambda calculus}
\end{figure}

@@

\section{Linear Haskell\label{linear-haskell-section}}

>>>>

\begin{itemize}
    \item Linear types were only added to the language roughly \emph{31 years
        after} Haskell's inception, unlike Rust and Idris2 which were
        designed with linearity from the start. It is an especially difficult
        endeavour to add linear types to a well-established language with a
        large library ecosystem and many active users, rather than to develop
        the language from the ground up with linear types in mind, and the
        successful approach as implemented in GHC 9.0, the leading Haskell
        compiler, was based on Linear Haskell\cite{linearhaskell}, where a
        linear type system designed with retaining backwards-compatibility and
        allowing code reuse across linear and non-linear users of the same
        libraries in mind was described. We describe Linear Haskell in detail in
        section~\ref{sec:linear-haskell}.

@@

In order to further give well defined semantics to a linear type system, we
present a linearly typed lambda calculus, a very simple language with linear
types, by defining what are syntatically valid programs through the grammar
in Fig.~\ref{fig:llcgrammar} and what programs are well typed through the typing
rules in Fig.~\ref{fig:llcrules}. The language features functions and function application ($\lolli$), two
flavours of pairs, additive ($\with$) and multiplicative ($\tensor$), a
disjunction operator ($\oplus$) and the $\bang$ modality operator. A typing
judgement for the linearly typed lambda calculus has the form
%
\[ \Gamma; \Delta \vdash M : A \]
%
where $\Gamma$ is the context of resources that may be used unrestrictedly, that
is, any number of times, $\Delta$ is the context of resources that must be used
linearly (\emph{exactly once}), $M$ is the program to type and $A$ is its type.
When resources from the linear context are used, they are removed from the
context and no longer available, and all resources in the linear context must be
used exactly once.

% Explicar o que é modality operator?

\begin{figure}[h]
\[
\begin{array}{c}
  \begin{array}{lcl}
    A,B & ::= & \star\\
        & \mid & A \lolli B\\
        & \mid & A \oplus B\\
        & \mid & A \tensor B\\
        & \mid & A \with B\\
        & \mid & \bang A\\
   \end{array}
\qquad
  \begin{array}{lcl}
      M,N & ::= & \star \mid \llet{\star = M}{N} \\
        & \mid & u \\
        & \mid & \lambda u. M \mid M~N\\
        & \mid & \textrm{inl}~M \mid \textrm{inr}~M \mid
        \ccase{M}{\textrm{inl}~u_1 \to N_1;\textrm{inr}~u_2 \to N_2}\\
        & \mid & M \tensor N \mid \llet{u_1 \tensor u_2 = M}{N}\\
        & \mid & M \with N \mid \textrm{fst}~M \mid \textrm{snd}~M \\
        & \mid & \bang M \mid \llet{!u = M}{N} \\
   \end{array}
\end{array}
\]
\caption{Grammar for a linearly-typed lambda calculus}
\label{fig:llcgrammar}
\end{figure}

The function abstraction is typed according to the linear function introduction
rule ($\lolli I$). The rule states that a a function abstraction, written
$\lambda u. M$, is a linear function (i.e. has type $A \lolli B$) given the
unrestricted context $\Gamma$ and the linear context $\Delta$, if the program $M$
has type $B$ given the same unrestricted context $\Gamma$ and the linear context
$\Delta,u{:}A$. That is, if $M$ has type $B$ using $u$ of type $A$ exactly once
besides the other resources in $\Delta$, then the lambda abstraction has the
linear function type.
%
\[
    \infer*[right=($\lolli I$)]
    {\Gamma ; \Delta , u{:}A \vdash M : B}
    {\Gamma ; \Delta \vdash \lambda u. M : A \lolli B}
\qquad
    \infer*[right=($\lolli E$)]
    {\Gamma ; \Delta \vdash M : A \lolli B \and \Gamma ; \Delta' \vdash N : A}
    {\Gamma ; \Delta, \Delta' \vdash M~N : B}
\]
%
Function application is typed according to the elimination
rule for the same type ($\lolli E$). To type an application $M~N$ as $B$, $M$
must have type $A \lolli B$ and $N$ must have type $A$. To account for the
linear resources that might be used while proving both that $M{:}A \lolli B$ and
$N{:}A$, the linear context must be split in two such that both typing judgments
succeed using exactly once every resource in their linear context (while the
resources in $\Gamma$ might be used unrestrictedly), hence the separation of the
linear context in $\Delta$ and $\Delta'$.

The multiplicative pair ($M \tensor N$) is constructed from two linearly typed expressions that
can each be typed with a division of the given linear context, as we see in its
introduction rule ($\tensor I$). Upon deconstruction, the multiplicative pair
elimination rule ($\tensor E$) requires that both of the pair constituents be
consumed exactly once.
%
\[
    \infer*[right=($\tensor I$)]
    {\Gamma ; \Delta \vdash M : A \and \Gamma ; \Delta' \vdash N : B}
    {\Gamma ; \Delta , \Delta' \vdash (M \tensor N) : A \tensor B}
\qquad
    \infer*[right=($\tensor E$)]
    {\Gamma ; \Delta \vdash M : A \tensor B \and \Gamma ; \Delta', u{:}A, v{:}B \vdash N : C}
    {\Gamma ; \Delta , \Delta' \vdash \llet{u \tensor v}{N} : C}
\]
%
On the other hand, the additive pair requires that both elements of the pair can
be proved with the same linear context, and upon deconstruction only one of the
pair elements might be used, rather than both simultaneously.

Finally, the modality operator ($\bang$) constructs a resource that can be used
unrestrictedly. Its introduction rule ($\bang I$) states that to construct this
resource means to add a resource to the unrestricted context, which can then be
used freely. To construct an unrestricted value, however, the linear context
must be empty -- an unrestricted value can only be constructed if it does not
depend on any linear resource.
%
\[
    \infer*[right=($\bang I$)]
    {\Gamma ; \cdot \vdash M : A}
    {\Gamma ; \cdot \vdash !M : !A}
\qquad    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta \vdash M : \bang A \and \Gamma, u{:}A ; \Delta' \vdash N : C}
    {\Gamma ; \Delta, \Delta' \vdash \llet{!u = M}{N} : C}
\]
%
To utilize an unrestricted value $M$, we must
bind it to $u$ with $\llet{\bang u = M}{N}$ and can then use it in $N$
unrestrictedly, because $u$ extends the unrestricted context rather than the
linear context as we have seen thus far.

In section~\ref{sec:linear-haskell} we describe how linear types are defined in
Haskell, a programming language more featureful than the linearly typed lambda
calculus. We will see that, the theoretical principles underlying the linear
lambda calculus and linear Haskell are the same, and by studying them in this
minimal setting we can understand them at large.

% TODO: Variable rules

\begin{figure}[h]
\[
  \begin{array}{c}
    \infer*[right=($\tensor I$)]
    {\Gamma ; \Delta \vdash M : A \and \Gamma ; \Delta' \vdash N : B}
    {\Gamma ; \Delta , \Delta' \vdash (M \tensor N) : A \tensor B}
\qquad
    \infer*[right=($\tensor E$)]
    {\Gamma ; \Delta \vdash M : A \tensor B \and \Gamma ; \Delta', u{:}A, v{:}B \vdash N : C}
    {\Gamma ; \Delta , \Delta' \vdash \llet{u \tensor v}{N} : C}
\\[1em]
    \infer*[right=($\lolli I$)]
    {\Gamma ; \Delta , u{:}A \vdash M : B}
    {\Gamma ; \Delta \vdash \lambda u. M : A \lolli B}
\qquad
    \infer*[right=($\lolli E$)]
    {\Gamma ; \Delta \vdash M : A \lolli B \and \Gamma ; \Delta' \vdash N : A}
    {\Gamma ; \Delta, \Delta' \vdash M~N : B}
\\[1em]
    \infer*[right=($\with I$)]
    {\Gamma ; \Delta \vdash M : A \and \Gamma ; \Delta \vdash N : B}
    {\Gamma ; \Delta \vdash M \with N : A \with B}
    \qquad
    \infer*[right=($\with E_L$)]
    {\Gamma ; \Delta \vdash M : A \with B}
    {\Gamma ; \Delta \vdash \textrm{fst}~M : A}
\qquad
    \infer*[right=($\with E_R$)]
    {\Gamma ; \Delta \vdash M : A \with B}
    {\Gamma ; \Delta \vdash \textrm{snd}~M : B}
\\[1em]
    \infer*[right=($\oplus I_L$)]
    {\Gamma ; \Delta \vdash M : A}
    {\Gamma ; \Delta \vdash \textrm{inl}~M : A \oplus B}
\qquad
    \infer*[right=($\oplus I_R$)]
    {\Gamma ; \Delta \vdash M : B}
    {\Gamma ; \Delta \vdash \textrm{inr}~M : A \oplus B}
\\[1em]
    \infer*[right=($\oplus E$)]
    {\Gamma ; \Delta \vdash M : A \oplus B \and \Gamma ; \Delta', w_1{:}A \vdash N_1 : C \and
    \Gamma ; \Delta', w_2{:}B \vdash N_2 : C}
    {\Gamma ; \Delta, \Delta' \vdash \ccase{M}{\textrm{inl}~w_1 \to N_1 \mid \textrm{inr}~w_2 \to N_2} : C}
    \\[1em]
    \infer*[right=($\star I$)]
    { }
    {\Gamma ; \cdot \vdash \star : \star}
\quad    \infer*[right=($\star E$)]
    {\Gamma ; \Delta \vdash M : \star \and \Gamma ; \Delta' \vdash N : B}
    {\Gamma ; \Delta, \Delta' \vdash \llet{\star = M}{N} : B}
\\[1em]
    \infer*[right=($\bang I$)]
    {\Gamma ; \cdot \vdash M : A}
    {\Gamma ; \cdot \vdash !M : !A}
\qquad    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta \vdash M : \bang A \and \Gamma, u{:}A ; \Delta' \vdash N : C}
    {\Gamma ; \Delta, \Delta' \vdash \llet{!u = M}{N} : C}
  \end{array}
\]
\caption{Typing rules for a linearly-typed lambda calculus}
\label{fig:llcrules}
\end{figure}

@@

\section{Linear Haskell\label{sec:linear-haskell}}
